What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Some proofreading of Section 6

Looking closely at the value types in variable updates, we observe that the
majority of \eval sites involve basic R vectors (\SEBasicTypeRatio) and lists
(\SEListTypeRatio). From the compiler perspective, we would like to know how
many sites changes function bindings. In the corpus, this happens in only
\SEClosureType sites; \SEClosureTypeLocal of them do that in the local
environment. We have manually inspected a few of these sites, but except for
manual injection of parameters into a \c{model.frame} execution environment, we
did not find a common use case.

@@

\mypara{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere, 
making it \eval more dangerous than it already is. However, the data  suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript, \footnote{\citep{ecoop11} shows that in the \emph{Interactive} scenario, \eval in Javascript performs, store events can reach up to 40\% of the events, and 7\% to 8\% of the \eval do side effects in the global scope. } and that over half of them happen in a predictable environment.

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are the tools given to developers to
this end.
%This API is slightly more complex than just passing a string, it is
%conceivable that this may discourage some casual users. \Eval is also being
%used to reduce boilerplate code and provide convenience features for
%programmers. 
In this section we focus on the \emph{why}, giving real-world examples of the
\eval usage in the wild. All of the examples are based on code from our corpus
often simplified to fit the space and increase clarity. For each example, we
indicate the package and the function where it is located.

@@

% We start the discussion with the examples for the minimized expressions from
% Section~\ref{sec:minimized}.
In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
expressions passed to \eval. Here we illustrate them on concrete instances.

@@

  \item \emph{Variable lookup and values and indexing \c{(X, V, \$)}}
    constitute one of the most common use case for \eval. Let's consider the
    \c{subset} function from R base library which allows, among others, to
    select variables in a data frame. For example, given a data frame \c{df}
    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
    sel=1)}. However, thanks to \eval it can also be also called with
    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
    \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
    returns the expression passed to \c{sel}, the second is the data frame in
    which this expression will be evaluated and finally the last indicates
    which environment should be used to look up bindings not found in \c{nl}.

    Another example is the most often called \eval site in CRAN:
    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function
    in \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
    find the parent class. In the vast majority of cases the variable is symbol
    and so the \eval seems superfluous. The \eval is however necessary so one
    can specify parent classes using \c{pkg\_name::class\_name} as the \c{::}
    operator is a function call in R.

  \item \emph{Assignments \c{(<-)}} appears in the cases where a developer wants
    to create or update a biding in a specific environment where either the
    name of the value come from some expression. Most cases come from trivial
    code generation where the assign expression is assembled using \c{parse} or
    \c{substitute}, often in a loop: \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-", i)))|.

    There are also more sophisticated cases. The \c{plyr::allocate\_column}
    function uses \eval for a deferred assignment to fill missing values in a
    column. Package \c{overture} repeatedly evaluates R expressions saving any
    assignments that happen in it, as samples of Markov chains.

  \item \emph{Function definition \c{(FUN)}} is mostly used in conjunction with
    code generation where new functions are synthesized using either \c{parse}
    or \c{substitute}. Some FFI frameworks use this to automatically generate R
    binding to native code. For example, \c{Rcpp}, which bridges R with C++,
    uses \eval to generate R function for C++ methods:\\
    \lstinline|eval(substitute(function(...) .CppObj$M(...), list(M=...)), env)|.

  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represents two very
    frequent patterns: \emph{code generation} and \emph{call reflection}. There are numerous
    cases for code generation. One occurring usage is generating ad-hoc calls
    to libraries such as \c{ggplot2} when one needs to parameterize data or
    aesthetics selection from a variable. For example \lstinline|ggplot(df, aes(x=A, y=B))|
    maps X and Y axes in the plot to \c{A} and \c{B} columns in
    \c{df}. It might not be immediately clear\footnote{\cf
    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
    how modify the call so it will pick the actual mapping from \c{A} and \c{B}
    variables. Another case is the traditional use of \eval as a means to
    reduce boilerplate code; simple and repetitive code can easily be replaced
    with judicious use of \eval. For example, the \c{data.table} package uses
    \eval to calls the \c{options} function with named arguments taken from a
    vector of strings. While the benefits are limited in this example, it is an
    attractive tool for programmers.
    %
    \begin{lstlisting} 
    opts = c("datatable.verbose"="FALSE", # ...many others
    for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")"))) 
    \end{lstlisting}

    Call reflection refers to the combination of \eval with \c{match.call},
    function that returns the current call with all its arguments captures as
    expressions. This is used mostly in the following circumstances: (1) to
    record a call for a later use in possibly different environment, (2) pass
    most of the captured call arguments to another function or (3) evaluate
    arguments of the function in different environments. This is widely used in
    relation with statistical modeling in combination with a model fitting
    function or with \c{model.frame} and related functions. For example the
    \c{survival::coxph} function\footnote{Function fitting a Cox proportional hazards
    regression model.} contains:
    %
    \begin{lstlisting}
    Call <- match.call()
    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
    tform[[1L]] <- quote(stats::model.frame)  # change the function called
    mf <- eval(tform, parent.frame())
    \end{lstlisting}

@@

>>>>

Looking closely at the value types in variable updates, we observe that the
majority of \eval sites involve basic R vectors (\SEBasicTypeRatio) and lists
(\SEListTypeRatio). From the compiler perspective, we would like to know how
many sites change function bindings. In the corpus, this happens in only
\SEClosureType sites; \SEClosureTypeLocal of them do that in the local
environment. We have manually inspected a few of these sites, but except for
manual injection of parameters into a \c{model.frame} execution environment, we
did not find a common use case.

@@

\mypara{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope.
In R, given the support for first-class environments, it can happen anywhere, 
making it \eval more dangerous than it already is. However, the data  suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive} scenario, \eval in Javascript performs, store events can reach up to 40\% of the events, and 7\% to 8\% of the \eval do side effects in the global scope. } and that over half of them happen in a predictable environment.

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are the tools given to developers to
this end.
%This API is slightly more complex than just passing a string, it is
%conceivable that this may discourage some casual users. \Eval is also being
%used to reduce boilerplate code and provide convenience features for
%programmers. 
In this section, we focus on the \emph{why}, giving real-world examples of the
\eval usage in the wild. All the examples are based on code from our corpus,
often simplified to fit the space and increase clarity. For each example, we
indicate the package and the function where it is located.

@@

% We start the discussion with the examples for the minimized expressions from
% Section~\ref{sec:minimized}.
In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
expressions passed to \eval. Here, we illustrate them on concrete instances.

@@

  \item \emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
    constitute one of the most common use case for \eval. Let's consider the
    \c{subset} function from the R base library, which allows, among others, to
    select variables in a data frame. For example, given a data frame \c{df}
    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
    sel=1)}. However, thanks to \eval it can also be also called with
    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
    \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
    returns the expression passed to \c{sel}, the second is the data frame in
    which this expression will be evaluated, and the last indicates
    which environment should be used to look up bindings not found in \c{nl}.

    Another example is the most often called \eval site in CRAN:
    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function
    in the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
    find the parent class. In the vast majority of cases, the variable is a symbol
    and so the \eval seems superfluous. The \eval is however necessary so one
    can specify parent classes using \c{pkg\_name::class\_name}, as the \c{::}
    operator is a function call in R.

  \item \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
    to create or update a binding in a specific environment where either the
    name or the value come from some expression. Most cases originate from trivial
    code generation where the assign expression is assembled using \c{parse} or
    \c{substitute}, often in a loop: \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-", i)))|.

    There are also more sophisticated cases. The \c{plyr::allocate\_column}
    function uses \eval for a deferred assignment to fill missing values in a
    column. Package \c{overture} repeatedly evaluates R expressions capturing any  assignments that happen in it, as samples of Markov chains.

  \item \emph{Function definition \c{(FUN)}} is mostly used in conjunction with
    code generation where new functions are synthesized using either \c{parse}
    or \c{substitute}. Some FFI frameworks use this to automatically generate R
    binding to native code. For example, \c{Rcpp}, which bridges R with C++,
    uses \eval to generate R functions for C++ methods:\\
    \lstinline|eval(substitute(function(...) .CppObj$M(...), list(M=...)), env)|.

  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
    frequent patterns: \emph{code generation} and \emph{call reflection}. There are numerous
    cases for code generation. One occurring usage is to generate ad-hoc calls
    to libraries such as \c{ggplot2} when one needs to parameterize data or
    aesthetics selection from a variable. For example \lstinline|ggplot(df, aes(x=A, y=B))|
    maps X and Y axes in the plot to \c{A} and \c{B} columns in
    \c{df}. It might not be immediately clear\footnote{\cf
    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}  for a ggplot2 user 
    how to modify the call to pick the actual mapping from \c{A} and \c{B}
    variables. Another case is the traditional use of \eval as a means to
    reduce boilerplate code; simple and repetitive code can easily be replaced
    with judicious use of \eval. For example, the \c{data.table} package uses
    \eval to call the \c{options} function with named arguments taken from a
    vector of strings. While the benefits are limited in this example, it is an
    attractive tool for programmers.
    %
    \begin{lstlisting} 
    opts = c("datatable.verbose"="FALSE", # ...many others
    for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")"))) 
    \end{lstlisting}

    Call reflection refers to the combination of \eval with \c{match.call}, a
    function that returns the current call with all its arguments captured as
    expressions. This is used mostly in the following circumstances: (1) to
    record a call for a later use in a possibly different environment, (2) pass
    most of the captured call arguments to another function or (3) evaluate
    arguments of the function in different environments. This is widely used in
    relation with statistical modeling in combination with a model fitting
    function or with \c{model.frame} and related functions. For example the
    \c{survival::coxph} function\footnote{Function fitting a Cox proportional hazards
    regression model.} contains:
    %
    \begin{lstlisting}
    Call <- match.call()
    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
    tform[[1L]] <- quote(stats::model.frame)  # change the function called
    mf <- eval(tform, parent.frame())
    \end{lstlisting}

    \todo{Add that this pattern is advised by the R documentation of match.call}

@@

\todo{Say somewhere that the R community commonly calls call/argument capturing (with substitute or match.call), Non-Standard Evaluation.}
